---
title: "Environmental Justice Mapping"
author: Sandra Notaro
subtitle: Mapping Income and Race with Department of Environmental Conservation Sites in Erie, Niagara, and Cattaraugus Counties
---

# Introduction

Environmental justice mapping can be easily done with the EPA's EJSCREEN tool (https://ejscreen.epa.gov/mapper/). This mapping tool provides a great introduction to analyzing and comparing environmental justice issues, but if this were implemented in an upperclassmen college lab, this would not suffice for a three hour lab. Not only is mapping a useful skill, but understanding how to map in RStudio is very beneficial. The overall goal is to develop a step-by-step lab for students to create environmental justice maps in RStudio that are similar to EJSCREEN. The end product of these maps will thus be presented in this project.

Low-income and race are key indicators to observe when analyzing environmental justice. In the context of Western New York, how have these marginalized communities been impacted unequally compared to higher-income and white neighborhoods? How can students observe the hazardous environmental impacts marginalized communities face? Exploration and comparison of environmental injustice will help students understand the core inequalities issues in Western New York. Therefore, I hypothesize that the lower-income and black neighborhoods will live closer to sites that have been or are in the process of being remediated. Despite the fact that some sites will be remediated already, this is still an indicator that the site was once contaminated, so it historically shows contamination marginalized communities have lived near.

# Materials and methods

Packages
```{r packages, eval=T, echo=T, message=F, results='hide', cache = F}
library(tidycensus)
library(tidyverse)
library(tigris)
library(sf)
library(sp)
library(leaflet)
library(RColorBrewer)
library(scales)
library(kableExtra)
library(dplyr)
library(htmltools)
```

Accessing Census Data
```{r acs, eval=T, echo=T, results='hide', message=F, cache = F}
loadacs5 <- load_variables(2018, "acs5", cache = TRUE)
#api_key <- census_api_key("type_your_API_here")
# Request an API from here https://api.census.gov/data/key_signup.html

tract_income <- get_acs(geography = "tract", variables = "B19125_001", state = "NY", 
                        county = c("Erie", "Niagara", "Cattaraugus"), 
                        year = 2018, geometry = TRUE)
#head(tract_income)
st_tract_income <- tract_income %>%
  st_transform("+proj=longlat +datum=WGS84")

tract_race <- get_acs(geography = "tract", variables = "B02001_003", state = "NY",
                      county = c("Erie", "Niagara", "Cattaraugus"), 
                      year = 2018, geometry = TRUE)
#head(tract_race)
st_tract_race <- tract_race %>%
  st_transform("+proj=longlat +datum=WGS84")

race_eth <- c(White = "B03002_003", Black = "B03002_004", Indigenous = "B03002_005",
          Asian = "B03002_006", Hispanic = 	"B03002_012")
race_eth_acs <- get_acs(geography = "tract", state = "NY",
                      county = c("Erie", "Niagara", "Cattaraugus"), variables = race_eth,
                      summary_var = "B03002_001",
                      year = 2018, geometry = TRUE)
#head(race_eth_acs)
```

Gathering and Tidying the DEC Data. 
When filtering for 02 and A, this was to find the active remediation sites. Sites classified as 02 means the disposal of hazardous waste represents a significant threat to the environment or to health, or hazardous waste disposal has not been confirmed, but the site has been listed on the Federal National Priorities List. Sites classified as A means the site is active. Sites classified as 01 is another classification for sites that have contamination that constitutes a significant threat to public health and the environment, and the significant threat to public health and the environment is causing or presents an imminent danger of causing, irreversible or irreparable damage to the environment. Currently, this dataset does not include this classification code.
When filtering for 04, 05, and C, this was a way to find sites that have been remediated (but may not be up to environmentally safe standards). Sites classified as 04 means the site has been properly closed but still requires continued site management consisting of operation, maintenance and/or monitoring. These sites have not been brought into compliance with standards, criteria, or guidance. 05 means that the site has been properly closed and requires no continued managment. Sites classified as C means that the remediation has been satisfactorily completed.
```{r gather_dec, eval=T, echo=T, message=F, results='hide', cache = F}
DEC_data <- read.csv("https://data.ny.gov/api/views/c6ci-rzpg/rows.csv?accessType=DOWNLOAD&bom=true&format=true&sorting=true")
#View(dec_data)

hazardous_sites <- DEC_data %>%
  filter(Site.Class %in% c("02", "A") & County %in% c("Erie", "Niagara", "Cattaraugus"))
hazardous_sites_sf <- st_as_sf(hazardous_sites, coords = c("Longitude", "Latitude"), dim = "XY")

remediated_sites <- DEC_data %>%
  filter(Site.Class %in% c("04", "05", "C") & County %in% c("Erie", "Niagara", "Cattaraugus"))
remediated_sites_sf <- st_as_sf(remediated_sites, 
                                coords = c("Longitude", "Latitude"), 
                                dim = "XY")
```


Preparing the Map
```{r mapprep, eval=T, echo=F, results='hide', cache = F}
ny_tracts <- tracts(state = "NY", county = c("Erie", "Niagara", "Cattaraugus"), cb = TRUE) %>%
  st_transform("+proj=longlat +datum=WGS84")
ny_water <- area_water(state = "NY", county = c("Erie", "Niagara", "Cattaraugus")) %>%
  st_transform("+proj=longlat +datum=WGS84")
ny_roads <- primary_secondary_roads(state = "NY", year = 2018) %>%
  st_transform("+proj=longlat +datum=WGS84")
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
```


```{r, echo = F, results='asis'}
#data %>% 
  #slice(1:10) %>% #show only 1:n rows
  #kable(digits=2,align="c")%>% #make table and round to two digits
  #kable_styling(bootstrap_options = 
                  #c("striped", "hover", "condensed", "responsive")) #apply other formatting
```

# Results

Finding the Number of Tracts Where a Race or Ethnicity is the Largest for that Tract
```{r tally, message=F, warning=F, cache=F}
race_eth_lgst <- race_eth_acs %>%
  group_by(GEOID) %>%
  filter(estimate == max(estimate)) %>%
  select(NAME, variable, estimate) %>%
  st_set_geometry(NULL)

race_eth_summarize <- race_eth_lgst %>%
  group_by(variable) %>%
  tally() %>%
  rename(c("Race/Ethnicity" = variable, "Number of Tracts" = n))

kable(race_eth_summarize) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

Mapping Median Household Income Estimate in the Past 12 Months in Erie Niagara and Cattaraugus Counties


Mapping Estimate Total Black Population in Erie Niagara and Cattaraugus Counties
```{r demographic, eval=T, echo=T, cache=F}
income_pal <- colorBin(palette = "Greens", bins = 6, domain = st_tract_income$estimate)
race_pal <- colorBin(palette = "YlGnBu", bins = 5, domain = st_tract_race$estimate)

demographics_map <- leaflet() %>%
  setView(lng = -78.878738, lat = 42.880230, zoom = 7) %>%
  setMaxBounds(lng1 = -78.878738, lat1 = 45.027500, lng2 = -78.878738, lat2 = 40.440624) %>%
  addProviderTiles("Esri") %>% 
  addPolygons(data = st_tract_income, 
              weight = 2, 
              fillOpacity = 0.8, 
              color = ~ income_pal(estimate), 
              label = ~paste0("Estimate Household Income: ", dollar(estimate)), 
              group = "Income") %>%
  addLegend(position = "bottomleft", 
            pal = income_pal , 
            values = st_tract_income$estimate,
            title = "Median Household Income", opacity = 0.7,
            labFormat = labelFormat(prefix = "$")) %>%
  addPolygons(data = st_tract_race, 
              weight = 2, 
              fillOpacity = 0.8, 
              color = ~ race_pal(estimate), 
              label = ~paste0("Estimate Population: ", estimate), group = "Race") %>%
  addLegend(position = "bottomright", 
            pal = race_pal , 
            values = st_tract_race$estimate,
            title = "Black Population", 
            opacity = 0.7) %>%
  addLayersControl(overlayGroups = c("Income", "Race"))

demographics_map
```

```{r decmaps, eval=T, echo=T, cache=F}
dec_map <- leaflet() %>%
  setView(lng = -78.878738, lat = 42.880230, zoom = 7) %>%
  setMaxBounds(lng1 = -78.878738, lat1 = 45.027500, lng2 = -78.878738, lat2 = 40.440624) %>%
  addProviderTiles("Esri") %>%
  addCircleMarkers(data = hazardous_sites_sf, radius = 1.25, color = "red", 
                   popup = ~paste0("<b>", "Site Name: ", Program.Facility.Name, "</b>", 
                         "<br/>", "Address: ", Address1,
                         "<br/>", "Site Class: ", Site.Class),
                   group = "Hazardous Sites") %>%
  addCircleMarkers(data = remediated_sites_sf, radius = 1.25, color = "purple",
                   popup = ~paste0("<b>", "Site Name: ", Program.Facility.Name, "</b>", 
                         "<br/>", "Address: ", Address1,
                         "<br/>", "Site Class: ", Site.Class),
                   group = "Remediated Sites") %>%
  addLayersControl(overlayGroups = c("Hazardous Sites", "Remediated Sites"))
dec_map
```

# Conclusions

[~200 words]

Clear summary adequately describing the results and putting them in context. Discussion of further questions and ways to continue investigation.

# References

All sources are cited in a consistent manner